Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-06 paper_jasa.tex
re-worked the section on continuous covariate

After many small pilot studies with local personnel, it was clear that some care was needed to set up the human subjects experiments. It was best for a subject to see a block of 10 lineups with varying difficulty, with a reasonable number of ``easy'' lineups. The explanations about each experiment (below) includes and explanation of how the lineups were sampled and provided to the observers.

@@

The experiment is designed to study the ability of human observers to detect the effect of a single variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical UMP test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making a side-by-side boxplot of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and these were plotted in the same way as side-by-side boxplots. The actual data plot was randomly placed among these null data plots to produce the lineup. Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 11 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

@@

This experiment is designed for the same regression model (Equation \ref{multi}) with $p=1$ and covariate $X_1$ being continuous. $X_1$ is simulated from standard normal(0,1). Data is simulated with two sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The different values for slope parameter that are used to generate data are shown in Table \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ while generating actual data. For each combination of parameter, we generate at least three different actual data sets constituting a total of 65 lineups. For some parameter combinations we have more than three replications. Unlike what we did in section \ref{sec:category} we did not use any blocking structure to simulate actual data and null plots.

{\bf Steps to obtain lineup}
\begin{enumerate}
	\item For a given parameter setting obtain actual data set
	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
	\item Generate actual data plot (a scatter plot of $Y$ vs $X_1$ a least square line overlaid)
	\item Generate 19 null data sets from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and obtain 19 null plots
	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
\end{enumerate}

A lineup for this experiment is shown in figure \ref{fig:test_continuous}.  This lineup is generated using sample size ($n$)=100, slope($\beta$)=1.25 and and error SD($\sigma$)=5. The actual data plot location is 11.

@@

{\bf Presenting lineups for evaluation} Each subject got at least 10 lineups for evaluation. For each combination of sample size ($n$) and error SD ($\sigma$) we define the difficulty of the lineup based on the five slope ($\beta$) parameters we choose as shown in table \ref{tbl:experiment_params}. For the smallest slope parameter the difficulty is 0 and for the largest parameter the difficulty is 4. The 10 lineups are chosen as follows;

\begin{enumerate}
	\item For one combination of  sample size ($n$) and error SD ($\sigma$) randomly select five lineups with one from each difficulty level
	\item For another combination of sample size ($n$) and error SD ($\sigma$) select four lineups with one from each difficulty level 0 through 3
	\item Finally select another lineup with difficulty level 0
\end{enumerate}

>>>>

After many small pilot studies with local personnel, it was clear that some care was needed to set up the human subjects experiments. It was best for a subject to see a block of 10 lineups with varying difficulty, with a reasonable number of ``easy'' lineups. The explanations about each experiment (below) includes an explanation of how the lineups were sampled and provided to the observers, for each of the three experiments.

\green{Something probably needs to be said about subjects who do multiple tests, what checks are done.}

@@

The experiment is designed to study the ability of human observers to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical UMP test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making a side-by-side boxplot of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and these were plotted in the same way as side-by-side boxplots. The actual data plot was randomly placed among these null data plots to produce the lineup. Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 11 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

@@

This experiment is very similar to the previous one, except that there is a single continuous covariate and no second covariate (Equation \ref{multi} with $p=1$), following the test in Table \ref{tbl:stat_multiple}, case 2. Data is simulated with two sample sizes ($n=100, 300$), two standard deviations of the error ($\sigma=5, 12$), a variety of slopes ($\beta$), as given in Table  \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ and values for $X_1$ are simulated from $N(0,1)$. For each combination of parameters, three different actual data sets are produced, yielding a total of 65 lineups. \green{ I don't understand the next two sentences.} For some parameter combinations we have more than three replications. Unlike what we did in section \ref{sec:category} we did not use any blocking structure to simulate actual data and null plots.

The actual data plot is generated by making a scatterplot of $Y$ vs $X_1$ with the least squares regression line overlaid. To produce the null plots in the lineup null data was simulated from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and plotted using the same scatterplot method as the actual data. 
An example lineup from this experiment is shown in Figure \ref{fig:test_continuous}.  This lineup is generated using sample size ($n$)=100, slope($\beta$)=1.25 and and error SD($\sigma$)=5. The actual data plot location is 11. \green{Check this} This lineup, we think, is a difficult one. To select 10 lineups for a subject, each combination of sample size ($n$) and error SD ($\sigma$) is given a difficulty value based on the slope ($\beta$) parameters. For the smallest slopes the difficulty is 0 (hardest) and for the largest slopes the difficulty is 4 (easiest). For one combination of sample size and standard deviation, the subject sees a randomly selected lineup of each difficulty level, all the slopes, giving 5 lineups. Another four lineups are chosen from a second choice of sample size and standard deviation, with difficulty levels 0-3, and the last lineup shown was a randomly selected lineup with difficulty 0. The order that the lineups are shown to participants is randomized. \green{Is this a correct description of the procedure?} Each subject saw a substantial number of difficult lineups. The intention was noble, to get a better handle on the power of the lineups for small slopes, but practically this made is very tiring for the participants. It would be better to provide the subjects with a larger selection of easier lineups and just a few difficult ones. \green{perhaps should go into so much detail about this, or keep this for the conclusions section.}

%{\bf Steps to obtain lineup}
%\begin{enumerate}
%	\item For a given parameter setting obtain actual data set
%	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
%	\item Generate actual data plot (a scatter plot of $Y$ vs $X_1$ a least square line overlaid)
%	\item Generate 19 null data sets from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and obtain 19 null plots
%	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
%\end{enumerate}

@@

%{\bf Presenting lineups for evaluation} Each subject got at least 10 lineups for evaluation. For each combination of sample size ($n$) and error SD ($\sigma$) we define the difficulty of the lineup based on the five slope ($\beta$) parameters we choose as shown in table \ref{tbl:experiment_params}. For the smallest slope parameter the difficulty is 0 and for the largest parameter the difficulty is 4. The 10 lineups are chosen as follows;

%\begin{enumerate}
%	\item For one combination of  sample size ($n$) and error SD ($\sigma$) randomly select five lineups with one from each difficulty level
%	\item For another combination of sample size ($n$) and error SD ($\sigma$) select four lineups with one from each difficulty level 0 through 3
%	\item Finally select another lineup with difficulty level 0
%\end{enumerate}
